---
title: "EURO 2020 groupstage predictions: round of 16"
author: "Leonardo Egidi - DEAMS, University of Trieste, Italy. Mail: legidi@units.it"
date: "24 June 2021"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
  html_document:
    df_print: paged
    toc: yes
  include: null
  ioslides_presentation:
    highlight: tango
  prettydoc::html_pretty:
    css: style.css
    theme: leonids
    toc: yes
  beamer_presentation:
    highlight: tango
  slide_level: 2
  slidy_presentation:
    fig.height: 3
    fig.width: 4
    highlight: tango
graphics: yes
header-includes:
- \usepackage{color}
- \usepackage{bm}
institute: University of Trieste
fontsize: 10pt
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, fig.height =10, fig.width = 8, out.width='750px', dpi=200, 
  dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
library(MASS)
```
```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
```

# The statistical model (in brief)

We use a **double Poisson model with dynamic team-specific abilities** for the attack and the defence. Let $(X_{i}, Y_{i})$ denote the random number of goals scored by the home and the away team in the $i$-th game, $i=1,\ldots,n$, respectively. $\mathsf{ranking}$ denotes the Coca-Cola FIFA ranking at May 27th, 2021, whereas att and def denote the attack and the defence abilities, respectively.


\begin{align}
X_i| \lambda_{1i} &\sim \text{Poisson}(\lambda_{1i}),\\
Y_i|\lambda_{2i} &\sim \text{Poisson}(\lambda_{2i}),  \\
\log(\lambda_{1i}) &=\  \text{home} + \text{att}_{h_i, t}+ \text{def}_{a_i,t} + \frac{\gamma}{2}(\mathsf{ranking}_{h_i}-\mathsf{ranking}_{a_i}) \\
\log(\lambda_{2i}) & =\    \text{att}_{a_i,t} + \text{def}_{h_i,t} - \frac{\gamma}{2}(\mathsf{ranking}_{h_i}-\mathsf{ranking}_{a_i}), \ \ i=1,\ldots,n\ (\text{matches}), \\
\text{att}_{k, t} &\sim \ \mathcal{N}(\text{att}_{k, t-1}, \sigma^2), \\
\text{def}_{k, t} &\sim \  \mathcal{N}(\text{def}_{k, t-1}, \sigma^2),\\
\sum_{k=1}^{n_t} \text{att}_{k, }&=0, \  \sum_{k=1}^{n_t}\text{def}_{k, }=0, \ \ k=1,\ldots n_t \ (\text{teams}), \  t=1,\ldots, T \ (\text{times}).
\label{eq:scoring_rue}
\end{align}

Lines (1)-(2) display the likelihood's equations (two Poisson distributions); lines (3)-(4) display the log-linear models for the scoring rates $\lambda_{1}, \lambda_{2}$; lines (5)-(6) display the dynamic prior distributions for the attack and the defence parameters, respectively; line (7) displays the sum-to-zero identifiability constraints. Model fitting has been obtained through the Hamiltonian Monte Carlo sampling, 2000 iterations, 4 chains (```rstan``` package). The historical data used to fit the models come from: **Nations' League** (2019-2020), **Euro UEFA Qualifiers** (2020-2021), **World Cup UEFA Qualifiers** (2021), **UEFA Euro 2020** (groupstage matches).

The idea is to provide a dynamic predictive scenario: at the end of each match-day, the model will be refitted  to predict the remaining matches.

<!-- # Estimated attack/defence abilities -->

<!-- In the plot below we display the posterior intervals for the **attack** (red) and **defence** (blue) abilities estimated through the training set matches: the higher the attack and the lower the defence values for a given team, and the better is the estimated overall team's ability. -->

<!-- ```{r data, echo = FALSE, eval = TRUE, fig.height=26} -->
<!--  poisson -->
<!-- #foot_abilities(fit4, euro_data) -->
<!-- #log_lik_4 <- extract_log_lik(fit4) -->
<!-- #loo4 <- loo(fit4) -->


<!-- foot_abilities(fit4, euro_data) -->


<!-- ``` -->


# Groupstage predictions: round of 16 (26-29 June)

Posterior  matches probabilities from the posterior predictive distribution of the model above are displayed in the table below. **mlo** denotes the most likely exact outcome (in parenthesis, the corresponding posterior probability). Darker regions in the plots below denote more likely outcomes: on the $x$-axis the home goals, on the $y$-axis the away goals. 

```{r data2, echo = FALSE, out.width="80%", fig.width=11, eval = TRUE, fig.height=16}
library(tidyverse)
library(footBayes)
library(loo)

nations_data <- read.csv2(file="nations.csv", sep = ",")
euro_qual_data <- read.csv2(file="euro_qual.csv", sep = ",")
wc_qual_data <- read.csv2(file="wc.csv", sep = ",")
#View(nations_data)
#View(euro_qual_data)
#View(wc_qual_data)


nations_ristr <- nations_data  %>%
  select(date_GMT, home_team_name, away_team_name,
         home_team_goal_count, away_team_goal_count)
nations_ristr <- nations_ristr[-c(163, 164),]

euro_qual_ristr <- euro_qual_data  %>%
  select(date_GMT, home_team_name, away_team_name,
         home_team_goal_count, away_team_goal_count)

wc_qual_ristr <- wc_qual_data  %>%
  select(date_GMT, home_team_name, away_team_name,
         home_team_goal_count, away_team_goal_count)
wc_qual_ristr <- wc_qual_ristr[c(1:75),]

fifa_ranking <- read.csv2(file="ranking_fifa.csv", header=FALSE)
#fifa_ranking <- fifa_ranking[c(1:48),]
colnames(fifa_ranking) <- c("team_name", "ranking")




euro_data <- rbind(nations_ristr, euro_qual_ristr, wc_qual_ristr)
# redefining times from 1 to 12
times <- paste(substr(euro_data$date_GMT, 1, 3),
               substr(euro_data$date_GMT, 8, 11))
times <- as.factor(times)
levels(times) <- c(2, 1, 9, 5, 8, 4, 7, 3, 6)


euro_data$date_GMT <- as.numeric(as.vector(times))
euro_data <- arrange(euro_data, date_GMT)

# drop low teams

low_teams <- c("Faroe Islands", "Latvia",  "Malta",
               "San Marino", "Liechtenstein", "Gibraltar",
               "Andorra")

euro_data1<- euro_data %>% filter(is.na(match(home_team_name, low_teams)))
euro_data2<- euro_data1 %>% filter(is.na(match(away_team_name, low_teams)))
euro_data <- euro_data2
ngames_prev <- 12
euro_data_test <- data.frame(
                  date_GMT = rep(10, ngames_prev),
                  home_team_name = c("Italy", "Wales",
                                   "Denmark", "Russia",
                                   "England", "Austria",
                                   "Netherlands", "Scotland",
                                   "Poland", "Spain",
                                   "Hungary", "Germany"),

                  away_team_name = c("Turkey", "Switzerland",
                                   "Finland", "Belgium",
                                   "Croatia", "FYR Macedonia",
                                   "Ukraine", "Czech Republic",
                                   "Slovakia", "Sweden",
                                   "Portugal", "France"),

                  home_team_goal_count = c(3,1,0,0,1,3,3,0,1,0, 0,0),
                  away_team_goal_count = c(0,1,1,3,0,1,2,2,2,0,3,1))

euro_data <-rbind(euro_data, euro_data_test)

euro_data_test2 <- data.frame(
  date_GMT = rep(11, ngames_prev),
  home_team_name = c("Russia", "Turkey",
                      "Italy", "Ukraine",
                     "Denmark", "Netherlands",
                     "Sweden", "Croatia",
                     "England", "Hungary", 
                     "Germany", "Spain"),
  away_team_name = c("Finland", "Wales",
                     "Switzerland", "FYR Macedonia",
                      "Belgium", "Austria",
                     "Slovakia",  "Czech Republic",
                     "Scotland", "France",
                     "Portugal", "Poland"),
  
  home_team_goal_count = c(1,0,3,2,1,2,1,1,0,1,4,1),
  away_team_goal_count = c(0,2,0,1,2,0,0,1,0,1,2,1))

euro_data <-rbind(euro_data, euro_data_test2)

euro_data_test3 <- data.frame(
  date_GMT = rep(12, ngames_prev),
  home_team_name = c( "Italy","Switzerland",  
                      "Ukraine", "Netherlands",
                     "Denmark", "Finland", 
                     "Scotland","England",
                     "Sweden", "Spain",
                     "Portugal", "Germany" ),
  away_team_name = c("Wales", "Turkey",
                     "Austria", "FYR Macedonia",
                     "Russia", "Belgium", 
                     "Croatia", "Czech Republic",
                     "Poland","Slovakia",  
                    "France","Hungary"),
  
  home_team_goal_count = c(1,3,0,3,4,0,1,1,3,5,2,2),
  away_team_goal_count = c(0,1,1,0,1,2,3,0,2,0,2,2))

euro_data <-rbind(euro_data, euro_data_test3)

ngames_prev <- 8
euro_data_test_round16 <- data.frame(
  date_GMT = rep(13, ngames_prev),
  home_team_name = c("Wales", "Italy",
                     "Netherlands", "Belgium",
                     "Croatia", "France",
                     "England", "Sweden"),
  away_team_name = c( "Denmark", "Austria",
                      "Czech Republic", "Portugal",
                      "Spain", "Switzerland",
                      "Germany", "Ukraine"),
  home_team_goal_count = rep(NA, ngames_prev),
  away_team_goal_count = rep(NA, ngames_prev))

euro_data <-rbind(euro_data, euro_data_test_round16)


## dynamic models

fit4 <- stan_foot(data = euro_data,
                  model="double_pois",
                  iter = 2000, cores = 4, 
                  predict=ngames_prev,
                  ranking = fifa_ranking,
                  dynamic_type = "seasonal") # double

prob <- foot_prob(fit4, euro_data)
colnames(prob$prob_table) <- c("home", "away",
                               "home win", "draw", "away win", "mlo")
knitr::kable(prob$prob_table)
prob$prob_plot
```